Water discharge quantity has a variable impact on sediment transport capacity in subglacial channels

### 2023-09-26 Delaneyetal2023b.tex
daniel and andrews comments

\begin{keypoints}
\item Water discharge variations in subglacial channels are mainly accommodated by water velocity, not channel size.
\item Greater variability in water velocity causes sediment transport capacity to vary more in subglacial channels than in subaerial channels.
\item In interpreting sediment transport records, water discharge may represent processes including proglacial sediment mobilization and sediment access driving variations in sediment discharge.
\end{keypoints}

\begin{abstract}
  Sediment transport  in subglacial and subaerial channels depends on  water velocity and channel width, which controls the shear stress exerted on the channel bottom and the space over which to integrate it.
  In subaerial channels, depth, width, and velocity evolve quickly with water discharge, while in subglacial channels, water discharge changes mainly affect water velocity due to slow-evolving channel geometry.
  Parameterizations applied to hydrographs from an Alpine glacier and the Greenland Ice Sheet reveal greater variability in sediment transport capacity in subglacial channels.
  Furthermore, results show that high subglacial sediment transport rates can persist across various water discharges.
  Slowly evolving channel width accommodates some variability in subglacial channels.
  Despite differences in response to water discharge, sediment discharge measurements generally correlate better with subaerial hydraulics.
  This implies that other factors like sediment access or proglacial processes might be at play, affecting sediment transport capacity differently in subglacial and subaerial channels.
\end{abstract}

@@

Over long periods, processes such as glacier abrasion and quarrying sculpt landscapes and create sediment to be transported fluvially from under glaciers \cite<c.f.>{hallet1979,iverson2012,ugelvig2018}.
On shorter time periods, pressurized water transports sediment from underneath glaciers \cite{walder1994,creyts2013,beaud2018,delaney2019}, should enough sediment be present subglacially.

In a transport-limited regime, sediment discharge responds to the sediment transport capacity, defined as the amount of sediment the water can carry.
Sediment transport capacity depends on the shear stress between water and the sediment it flows  over \cite{shields1936,meyer1948,engelund1967} and the width of the channel bottom $w_c$ over which to mobilize sediment.
The shear stress  across the channel width $\tau_t$ responds to the velocity of water $v$ flowing through the channel so that
\begin{linenomath*}
  \begin{equation}
    \label{eq:tau_t}
    \tau_t \propto w_c\, v^2,
  \end{equation}
\end{linenomath*}
%
where $w_c$ is the width of the channel. Following mass conservation, the velocity of the water flowing through a  channel is 
\begin{linenomath*}
  \begin{equation}
    \label{eq:v}
    v = \frac{Q}{S},
  \end{equation}
\end{linenomath*}
where $Q$ is water discharge,  and $S$ is the channel's wetted area.

@@

The response of water velocity to changing water discharge in subglacial channels differs from subaerial ones.
The size of subglacial channels responds to the creep closure of the ice above and the opening of the channel by frictional heating of water flowing through the channel \cite{rothlisberger1972}.
As a result, the subglacial channel size only evolves relatively slowly over days to months, whereas water discharge can vary over hours \cite<e.g.>{iken1986,andrews2014,nanni2020}.
Therefore, subglacial water flow behaves more like pipe flow, and changes in water discharge $Q$ are mainly accommodated with increased or decreased water velocity $v$ \cite{alley1997}, because the size of the channel $S$ changes on a much longer timescale (Equation~\ref{eq:v} and Figure~\ref{fig:cartoon}).
It follows that an increase in water discharge $Q$ will cause the shear stress across the channel $\tau_t$ to respond with water velocity $v$, as opposed to subaerial channels where channel width $w_c$ evolves as well.

@@

As a result, sediment mobilization in subaerial and subglacial channels responds differently to changing water discharge.
These differences have been implicitly included in a wide range of available models that  quantify sediment transport from catchments in both subglacial and subaerial channels  \cite<e.g.>{walder1994,alley1997,tucker1997,creyts2013,delaney2019,hewitt2019,wickert2019}.
The divergent response of changing water discharge on sediment mobilization capacity may well impact sediment dynamics at glacier margins where flow transitions from pressurized to open channel flow \cite<e.g.>{lane2016,perolo2018}.
Additionally, the variable response of sediment transport capacity to water discharge in the two systems may affect the interpretation of sediment transport records in glacial systems \cite<e.g.>{muller1968,richards2003,swift2005,ganti2016}.
Yet, little work has examined the differing relationship between variations in water discharge and sediment discharge in subglacial and subaerial channel \cite{alley1997}.
%Furthermore, thorough observations of water velocity and channel width in subglacial channels are limited and, to these authors' knowledge, not contrasted with subaerial conditions at the glacier margins. %\mauro{self-cite here?}

@@

Water and suspended sediment data are leveraged from an Alpine and an ice sheet setting.
The Alpine site (\alpine) is  Fieschergletscher in the Swiss Alps ($46^\circ\,29'\,07''$ N, $8^\circ\,08'\,3''$ E).
Water discharge and suspended sediment concentration were collected here at a $1$\unit{min} interval continuously over the period 2014--2021.
See \citeA{felix2022} and \citeA{felix2021} for more details.

The Leverett Glacier in Greenland (\icesheet) serves as the ice sheet setting.
Water discharge and suspended sediment concentration data were collected roughly $2$\unit{km} downstream from the terminus from June to August/September 2009 to 2012 ($67^\circ\,03'\,5''$ N, $50^\circ\,12'\,59''$ W).
Data is used from \citeA{tedstone2017} and \citeA{hawkings2014} at a $5$\,\unit{min} time interval from May 28, 2012 to August 8, 2012.
Two gaps in suspended sediment data of roughly one day in length are interpolated across with a linear spline.

@@

Here, it is assumed that the water is transported through a subglacial channel flowing down the glacier; \cite<Figure~\ref{fig:cartoon}; >{rothlisberger1972}, and that the channel  size responds to frictional heating from water flow and creep closure by ice, using the Darcy-Weisbach formulation for water-flow through a pipe  \cite<e.g.>{rothlisberger1972,clarke2003}.
The formulation here does not consider the englacial storage of water.
Thus, it is assumed that the changing hydraulic head at the top of the glacier is negligible in the channel size evolution.
The evolution of subglacial channel size $S_g$ is then given as
\begin{linenomath*}
  \begin{equation}
    \label{eq:dS_dt}
    \frac{\partial S_g}{\partial t} = C_1 \frac{Q \Delta h}{l} - C_2 \left(h_{o}-\frac{\Delta h}{2}\right)^n\,S_g,
  \end{equation}
\end{linenomath*}
\noindent where $C_1= (1-\rho_wc_pc_t)\,\frac{\rho_wg}{\rho_iL}$ and $C_2=2A(\frac{\rho_wg}{n})^n$ are constants (individual quantities defined in Table \ref{table:vpm}), $l$ is the length of the subglacial channel, $Q$ is water discharge, $\Delta h$ is the hydraulic potential change from the glacier terminus to its top, $h_{o}= \frac{\rho_i}{\rho_w} h_{ice}$ is the ice overburden pressure, and $n$ is the exponent in Glen's law for ice flow \cite{glen1955} (usually $n=3$).
The first term on the right side of the equation represents the opening of the channel through frictional heating, while the following term represents the creep closure of the channel from ice deformation.

@@

The width-integrated shear stress can then be represented by
\begin{equation}
  \label{eq:tautg}
  \tau_{tg}=w_{cg}\,\tau_g,
\end{equation}

@@

To parameterize the shear stress of water flowing across sediments in the subaerial channel,  the hydraulics parameterization presented in \citeA{tucker1997} is implemented.
Using mass conservation and the Darcy-Weisbach relationship as well as assuming that the channel is wide so that the hydraulic radius is well approximated by the flow depth, then
the shear stress $\tau_f$ at the river bed is
\begin{linenomath*}
  \begin{equation}
    \label{eq:DW_tau}
    \tau_f=\frac{\rho_w\,g^{\frac{2}{3}}\,f_p^{\frac{1}{3}}}{2}\, \Big(\frac{Q}{w_{cf}} \Big)^{\frac{2}{3}} \,\nabla z_c^{\frac{2}{3}},
  \end{equation}
\end{linenomath*}
where $\nabla z_c$ is the channel slope, and $f_p$ is the friction factor for subaerial channels.
Channel width $w_{cf}$ is
\begin{equation}
  \label{eq:wcf}
  w_{cf} = k \, Q^\frac{1}{2},
\end{equation}
%
where $k$ is a constant and $\frac{1}{2}$ is a commonly chosen exponent \cite{leopold1953}.
%\mauro{This is for so called ``regime channels''}
Water velocity, $v_f$, is given by rearranging Equation \ref{eq:tau} as
\begin{equation}
  \label{eq:vf}
  v_f = \sqrt{\frac{8\,\tau_f}{f_p\,\rho_w}}.
\end{equation}
%
As above, the width-integrated shear stress is then
\begin{equation}
    \label{eq:tautf}
    \tau_{tf}=w_{cf}\,\tau_f.
  \end{equation}

@@

For each glacier, the parameterizations are first applied to a reference test case that assumes  subglacial channel with $\beta=\frac{\pi}{2}$. In this case the friction factors $f_r$ and $f_p$ are tuned so that the parameterizations reproduce reasonable velocities for both  the Greenland Ice Sheet and the Alpine glacier \cite<$\sim\,1.6\,$\unit{m}$^3$\,\unit{s}$^{-1}$>{werder2010b,chandler2013}.
In both subaerial channels, the bed slope is $\nabla z_c = 0.05$.
Water velocity, shear stress, and width-integrated shear stress in the subglacial and subaerial channels from this formulation are presented below.

Second, to characterize the variability in sediment discharge capacity in subglacial compared to subaerial channels further, $100$  parameterizations are applied to a range of different channel geometry factors ($\beta$ and $k$) and friction factors ($f_r$ and $f_p$).
Runs with parameter combinations are culled if their mean subglacial water velocity over the season lies outside of the range between $0.75$\,\unit{m}$^3$\,\unit{s}$^{-1}$ and $2$\,\unit{m}$^3$\,\unit{s}$^{-1}$ or if subaerial water velocity lies outside of the range $0.3$\,\unit{m}$^3$\,\unit{s}$^{-1}$ and $1.2$\,\unit{m}$^3$\,\unit{s}$^{-1}$, in accordance with measurements \cite<e.g.>{werder2010b,magnusson2012,chandler2013}.
For any given variable, the standard deviation of that variable over the course of the season is used to quantify its variability.

Lastly,  outputs are compared with the sediment discharge records of the glaciers using Spearman rank correlation, that is the correlation with respect to the ordering of values from the outputs and the observations.
Rank correlation reduces the impact of the non-linear response of sediment transport capacity to the hydrological forcing.
Periods of uncertainty in the data in \icesheet are removed from the analysis.
Outputs and observations, which have a native time resolution of $15$\,\unit{min}, are aggregated across timescales from $1$\,\unit{hr} to $15$\,\unit{days} by averaging values one half of the timescale before and after a given point time.

@@

Over the course of the season, large peaks in water velocity and shear stress, along with width-integrated shear stress, occur in response to increases in water discharge in both \alpine{} and \icesheet{}.
This occurs as the subglacial values are generally larger than the subaerial values, thus more efficient at transporting sediment.
In the subglacial channel, peaks generally occur over the period when water discharge increases at the fastest rate (Figure~\ref{fig:model_outs}).
Alternatively, in the subaerial channel, peaks in these sediment transport variables occur when water discharge is highest.
The timing of the greatest rate of change in water discharge occurs before the peaks in water discharge.
Here, sediment could be deposited in the proglacial area as the water depressurizes, then remobilized when sediment transport capacity peaks in the subaerial system.

@@

Each water discharge value in the subaerial channel results in a unique water velocity, shear stress, and width-integrated shear stress, creating a perfect rank correlation between the variables (Figure~\ref{fig:Qw_vari} a--f).
In the subglacial channel, however,  water velocity, shear stress, and width-integrated shear stress can vary substantially for a given water discharge and width-integrated shear stress varies with respect to the other two  outputs as well (Figure~\ref{fig:Qw_vari} a--f).
For instance, very high  water velocity values and shear stresses can occur across minimal water discharge at $2.5$\,\unit{m}$^3$\,\unit{s}$^{-1}$ to the maximum water discharge at over $17$ \,\unit{m}$^3$\,\unit{s}$^{-1}$.
In \icesheet, mean values of water velocity can occur at water discharges between roughly $150$ \,\unit{m}$^3$\,\unit{s}$^{-1}$ and $310$ \,\unit{m}$^3$\,\unit{s}$^{-1}$, effectively spanning much of the water discharge range (Figure~\ref{fig:Qw_vari} d--f).

When considering the subglacial channel's evolving width, width-integrated shear stress  across the channel generally increases with water discharge, with improved rank correlation compared to water discharge and water velocity or shear stress (Figure~\ref{fig:Qw_vari} a--f).
Yet even the width-integrated shear stress  can vary substantially, with the highest values occurring at water discharge values ranging from roughly $11$ \,\unit{m}$^3$\,\unit{s}$^{-1}$, to over $17$ \,\unit{m}$^3$\,\unit{s}$^{-1}$.
The variability in width-integrated shear stress is less pronounced in the \icesheet case, with reduced variability in water discharge (Figure~\ref{fig:Qw_vari} c, f).
Yet, a range of width-integrated shear stress can occur at a given discharge.

@@

The second application examines the impact of a variety of channel shapes and friction factors on the relationship between water discharge and sediment transport capacity in both systems.
Across the range of parameter values examined, variability in water velocity and shear stress remains higher in the subglacial system compared to the subaerial one.
However, in \alpine{}, the variability in width-integrated shear stress is comparable between the two systems, due to the variable water discharge (Figure~\ref{fig:multi_run}).
 However, in \icesheet{}, with lower water discharge variability, the variability in width-integrated shear stress remains higher in the subglacial system across the time aggregation periods.

It is prescient to evaluate the time scales over which the subglacial channels may respond to changing water discharge to understand the influence of variability in sediment transport capacity.
In both cases, the variability in velocity, shear stress, and width-integrated shear stress decrease after $1-5$ day aggregations (Figure~\ref{fig:multi_run}).
Yet, substantial differences in variability persist in periods of up to $15$ days, as the variability in aggregations periods longer than this might be conflated with seasonal variations in water discharge.
Note that lags and time of peak sediment transport conditions still occur over these periods (Figure~\ref{fig:model_outs}), while the variability in sediment transport parameters decreases with larger time aggregation (Figure~\ref{fig:multi_run}).

\subsection{Relationship between parameterization  outputs and observed sediment discharge}
When comparing the output variables with suspended sediment discharge measurements, the variables velocity, shear stress, and width-integrated shear stress from the subaerial system generally show better correlation with sediment discharge compared to subglacial outputs across the tested subglacial channel parameters (Figures~\ref{fig:Qw_vari} g--l and \ref{fig:multi_run}).
The rank correlation amongst the  subaerial parameterization's three output variables remains fairly constant despite the variable impact of channel shape and consistent scaling with water discharge (Figure \ref{fig:Qw_vari} and Section~\ref{sect:fluv}).
In subglacial channels, velocity and shear stress perform substantially worse in terms of rank correlation with sediment discharge compared to width-integrated shear stress (Figures~ \ref{fig:multi_run}).
In both \alpine{} and \icesheet{}, the  subglacial rank correlation approaches that of subaerial channels over longer time aggregation periods (Figures~ \ref{fig:multi_run} and \ref{fig:model_outs_1hr} to \ref{fig:model_outs_15day}).

\section{Discussion}
\subsection{Context of sediment capacity differences between the channel types}
\label{sect:dis_qsc}
Quantities that control sediment transport capacity, such as water velocity, shear stress, and width-integrated shear stress, vary differently in response to water discharge in subglacial channels compared to subaerial ones (Figure~\ref{fig:Qw_vari}).
In sediment transport relationships that evaluate sediment discharge capacity, such as in \citeA{meyer1948} or \citeA{engelund1967}, shear stress is scaled to the power of $\frac{3}{2}$ or $\frac{5}{2}$, respectively.
The exponent greater than $1$ magnifies sediment discharge variability beyond the variable sediment transport parameters described above (Figure~\ref{fig:multi_run}).
Note that sediment transport relationships describe sediment flux per unit width, so that the integrated capacity across channel occurs after application of the exponent, unlike the width-integrated shear stress evaluation here \cite<Equations\,\ref{eq:tautg} and \ref{eq:tautf}; c.f.>{tucker1997}.
Note also that both subglacial and subaerial parameterizations here assume that the distribution of velocity and shear stress is homogeneous across the channel bed, despite the limitations of this assumption \cite<e.g.>{yager2018}.

\ian{Mauro: Can you double check this?}
The rapid increases and decreases in subglacial sediment transport capacity may compound the variability in sediment transport already present in subaerial systems \cite<Figure~\ref{fig:Qw_vari}; >{williams1989,jerolmack2010}.
By applying the discharge - width scaling in Equation~\ref{eq:wcf} to Equation~\ref{eq:v}, by setting $S\, \propto \,Q^{\frac{1}{2}}$, the subaerial channel's width-integrated shear stress response is $\tau_{tf}\, \propto\,  Q$ (see \ref{sect:scaling}). 
For sufficiently short time periods,  channel width and wetted area are fixed in the subglacial system so that the response of width-integrated shear stress is $ \tau_{tg}\, \propto\,  Q^{2}$, with a larger exponent on $Q$ than in the subaerial case following \citeA{alley1997}.
In this case, water discharge fluctuates at a shorter timescale than the subglacial channel evolution \cite{gimbert2016}.
Note also that the subglacial width integrated shear stress response to water discharge will be greater than the subaerial's provided that  Equation~\ref{eq:wcf}'s exponent is below one.
However, when subglacial channel size lies in equilibrium with water discharge, the  width-integrated shear stress response is $\tau_{tg}\, \propto \, Q^{\frac{4}{5}}$, similar to subaerial channels.
It follows that the stronger correlation between hydraulics and water discharge in \icesheet{} could result from being closer to this equilibrium between variations in the channels' wetter area and water discharge compared to \alpine{}.
This state could result from the reduced diurnal discharge amplitude in \icesheet{} (Figures~\ref{fig:model_outs} and ~\ref{fig:Qw_vari}).
In both cases, hydraulics could trend toward this equilibrium regime when average discharge over $1-5$ days when variability in subglacial channels  approaches that of subaerial channels (Figure~\ref{fig:multi_run}).

>>>>

\begin{keypoints}
\item Water discharge variations in subglacial channels are mainly accommodated by water velocity, not channel size.
\item Greater variability in water velocity causes sediment transport capacity to vary more in subglacial channels than in subaerial channels.
\item Water discharge may represent processes including proglacial sediment mobilization and sediment access driving variations in sediment discharge, as opposed to subglacial sediment transport capacity in glacierized catchments.
\end{keypoints}

\begin{abstract}
  Sediment transport capacity in subglacial and subaerial channels depends on  water velocity and channel width, which together control the shear stress exerted on the channel bottom.
  In subaerial channels, width, and velocity evolve quickly with water discharge, while in subglacial channels, water discharge changes mainly affect water velocity due to slow-evolving channel geometry.
  Parameterizations of these subglacial and subaerial hydraulics are applied to hydrographs from an Alpine glacier and the Greenland Ice Sheet revealing greater variability in sediment transport capacity in subglacial channels.
  Furthermore, results show that high subglacial sediment transport rates can persist across various water discharges.
  Slowly evolving channel width accommodates some variability in subglacial channels.
  Despite differences in response to water discharge, sediment discharge measurements generally correlate better with subaerial hydraulic parameters.
  This implies that other factors like sediment access or proglacial processes might be at play, affecting sediment transport capacity differently in subglacial and subaerial channels.
\end{abstract}

@@

Over long periods, processes such as glacier abrasion and quarrying sculpt landscapes and create sediment to be transported fluvially from under glaciers \cite<c.f.>{hallet1979,iverson2012,ugelvig2018}.
On shorter time periods, pressurized water transports sediment from underneath glaciers \cite{walder1994,creyts2013,beaud2018,delaney2019}, should enough sediment be present subglacially (i.e. in a transport-limited regime).

In a transport-limited regime, sediment discharge responds to the sediment transport capacity, defined as the amount of sediment the water can carry.
Sediment transport capacity depends on the shear stress between water and the sediment it flows  over \cite{shields1936,meyer1948,engelund1967} along with the width of the channel bottom $w_c$ over which to mobilize sediment.
The shear stress  across the channel width $\tau_t$ responds to the velocity of water $v$ flowing through the channel so that
\begin{linenomath*}
  \begin{equation}
    \label{eq:tau_t}
    \tau_t \propto w_c\, v^2,
  \end{equation}
\end{linenomath*}
%
where $w_c$ is the width of the channel. Following mass conservation, the velocity of the water flowing through a  channel is 
\begin{linenomath*}
  \begin{equation}
    \label{eq:v}
    v = \frac{Q}{S},
  \end{equation}
\end{linenomath*}
where $Q$ is water discharge,  and $S$ is the channel's wetted area.

@@

The response of water velocity to changing water discharge in subglacial channels differs from subaerial ones.
The size of subglacial channels responds to the creep closure of the ice above and the opening of the channel by frictional heating of water flowing through the channel \cite{rothlisberger1972}.
As a result, the subglacial channel size only evolves relatively slowly over days to months, whereas water discharge can vary over hours \cite<e.g.>{iken1986,andrews2014,nanni2020}.
Therefore, subglacial water flow behaves more like pipe flow, and changes in water discharge $Q$ are mainly accommodated with increased or decreased water velocity $v$ \cite{alley1997} because the size of the channel $S$ changes on a much longer timescale (Equation~\ref{eq:v} and Figure~\ref{fig:cartoon}).
It follows that an increase in water discharge $Q$ will cause the shear stress across the channel $\tau_t$ to respond with water velocity $v$, as opposed to subaerial channels where channel width $w_c$ evolves as well.

@@

As a result, sediment mobilization in subaerial and subglacial channels responds differently to changing water discharge.
These differences have been implicitly included in a wide range of available models that  quantify sediment transport from catchments in both subglacial and subaerial channels  \cite<e.g.>{walder1994,alley1997,tucker1997,creyts2013,delaney2019,hewitt2019,wickert2019}.
The divergent response of changing water discharge on sediment mobilization capacity may well impact sediment dynamics at glacier margins where flow transitions from pressurized to open channel flow \cite<e.g.>{lane2016,perolo2018}.
Additionally, the variable response of sediment transport capacity to water discharge in the two systems may affect the interpretation of sediment transport records in glacial systems \cite<e.g.>{muller1968,richards2003,swift2005,ganti2016}.
Yet, little work has examined the differing relationship between variations in water discharge and sediment discharge in subglacial and subaerial channels \cite{alley1997}.
%Furthermore, thorough observations of water velocity and channel width in subglacial channels are limited and, to these authors' knowledge, not contrasted with subaerial conditions at the glacier margins. %\mauro{self-cite here?}

@@

Water and suspended sediment data are leveraged from Alpine and ice sheet settings.
The Alpine site (\alpine) is  Fieschergletscher in the Swiss Alps ($46^\circ\,29'\,07''$ N, $8^\circ\,08'\,3''$ E).
Water discharge and suspended sediment concentration were collected here at a $1$\unit{min} interval continuously over the period 2014--2021.
See \citeA{felix2022} and \citeA{felix2021} for more details.

The Leverett Glacier in Greenland (\icesheet) serves as the ice sheet setting.
Water discharge and suspended sediment concentration data were collected roughly $2$\unit{km} downstream from the terminus between June and August/September 2009 to 2012 ($67^\circ\,03'\,5''$ N, $50^\circ\,12'\,59''$ W).
Data is used from \citeA{tedstone2017} and \citeA{hawkings2014} at a $5$\,\unit{min} time interval from May 28, 2012 to August 8, 2012.
Two gaps in suspended sediment data of roughly one day in length are interpolated across with a linear spline.

@@

Here, it is assumed that the water is transported through a subglacial channel \cite<Figure~\ref{fig:cartoon}; >{rothlisberger1972}, and that the channel  size responds to frictional heating from water flow and creep closure by ice, using the Darcy-Weisbach formulation for water-flow through a pipe  \cite<e.g.>{rothlisberger1972,clarke2003}.
The formulation here does not consider the englacial storage of water.
Thus, it is assumed that the changing hydraulic head at the top of the glacier is negligible in the channel size evolution.
The evolution of subglacial channel size $S_g$ is given as
\begin{linenomath*}
  \begin{equation}
    \label{eq:dS_dt}
    \frac{\partial S_g}{\partial t} = C_1 \frac{Q \Delta h}{l} - C_2 \left(h_{o}-\frac{\Delta h}{2}\right)^n\,S_g,
  \end{equation}
\end{linenomath*}
\noindent where $C_1= (1-\rho_wc_pc_t)\,\frac{\rho_wg}{\rho_iL}$ and $C_2=2A(\frac{\rho_wg}{n})^n$ are constants (Table \ref{table:vpm}), $l$ is the length of the subglacial channel, $Q$ is water discharge, $\Delta h$ is the hydraulic potential change from the glacier terminus to its top, $h_{o}= \frac{\rho_i}{\rho_w} h_{ice}$ is the ice overburden pressure, and $n$ is Glen's n \cite{glen1955} (usually $n=3$).
The first term on the right side of the equation represents the opening of the channel through frictional heating, while the following term represents the creep closure of the channel from ice deformation.

@@

The width-integrated shear stress is represented as
\begin{equation}
  \label{eq:tautg}
  \tau_{tg}=w_{cg}\,\tau_g,
\end{equation}

@@

To parameterize the shear stress of water flowing across sediments in the subaerial channel,  the hydraulics parameterization presented in \citeA{tucker1997} is implemented.
Using mass conservation and the Darcy-Weisbach relationship as well as assuming that the channel is wide so that the hydraulic radius is well approximated by the flow depth, then
the shear stress $\tau_f$ at the river bed is
\begin{linenomath*}
  \begin{equation}
    \label{eq:DW_tau}
    \tau_f=\frac{\rho_w\,g^{\frac{2}{3}}\,f_p^{\frac{1}{3}}}{2}\, \Big(\frac{Q}{w_{cf}} \Big)^{\frac{2}{3}} \,\nabla z_c^{\frac{2}{3}},
  \end{equation}
\end{linenomath*}
where $\nabla z_c$ is the channel slope, and $f_p$ is the friction factor for subaerial channels.
Channel width $w_{cf}$ is
\begin{equation}
  \label{eq:wcf}
  w_{cf} = k \, Q^\frac{1}{2},
\end{equation}
%
where $k$ is a constant and $\frac{1}{2}$ is a commonly chosen exponent \cite{leopold1953}.
%\mauro{This is for so called ``regime channels''}
Water velocity, $v_f$, is given by rearranging Equation \ref{eq:tau} as
\begin{equation}
  \label{eq:vf}
  v_f = \sqrt{\frac{8\,\tau_f}{f_p\,\rho_w}}.
\end{equation}
%
As above, the width-integrated shear stress is
\begin{equation}
    \label{eq:tautf}
    \tau_{tf}=w_{cf}\,\tau_f.
  \end{equation}

@@

For each glacier, the parameterizations are first applied to a reference test case for each glacier that assumes  subglacial channel with $\beta=\frac{\pi}{2}$. Friction factors $f_r$ and $f_p$ are tuned so that the parameterizations reproduce reasonable water velocities for both \alpine and \icesheet{}\cite<$\sim\,1.6\,$\unit{m}$^3$\,\unit{s}$^{-1}$>{werder2010b,chandler2013}.
In both subaerial channels, the bed slope is $\nabla z_c = 0.05$.
Water velocity, shear stress, and width-integrated shear stress in the subglacial and subaerial channels from this formulation are presented below.

Second, to characterize the variability in sediment discharge capacity in subglacial compared to subaerial channels further, $100$  parameterizations are applied to a range of different channel geometry factors ($\beta$ and $k$) and friction factors ($f_r$ and $f_p$).
Runs with parameter combinations are culled if their mean subglacial water velocity over the season lies outside of the range between $0.75$\,\unit{m}$^3$\,\unit{s}$^{-1}$ and $2$\,\unit{m}$^3$\,\unit{s}$^{-1}$ or if subaerial water velocity lies outside of the range $0.3$\,\unit{m}$^3$\,\unit{s}$^{-1}$ and $1.2$\,\unit{m}$^3$\,\unit{s}$^{-1}$, in accordance with measurements \cite<e.g.>{werder2010b,magnusson2012,chandler2013}.
The standard deviation of an output over the course of the season is used to quantify its variability.

Lastly,  outputs are compared with the sediment discharge records of the glaciers using Spearman rank correlation, that is the correlation with respect to the ordering of values from the outputs and the observations.
Rank correlation reduces the impact of the non-linear response of sediment transport capacity to the hydrological forcing.
Outputs and observations, which have a native time resolution of $5$\,\unit{min}, are aggregated across timescales from $15$\,\unit{min} to $15$\,\unit{days} by averaging values one half of the timescale before and after a given point time.

@@

Over the course of the season, large peaks in water velocity and shear stress, along with width-integrated shear stress, occur in response to increases in water discharge in both \alpine{} and \icesheet{}.
This occurs as the subglacial outputs are generally larger than the subaerial values, thus more efficient at transporting sediment.
In the subglacial channel, peaks generally occur over the period when water discharge increases at the fastest rate (Figure~\ref{fig:model_outs}).
Alternatively, in the subaerial channel, peaks in these sediment transport variables occur when water discharge is highest.
The timing of the greatest rate of change in water discharge occurs before the peaks in water discharge.
Here, sediment could be deposited in the proglacial area as the water depressurizes, then remobilized when sediment transport capacity peaks in the subaerial system.

@@

Each water discharge value in the subaerial channel results in a unique water velocity, shear stress, and width-integrated shear stress, creating a perfect rank correlation between the variables (Figure~\ref{fig:Qw_vari} a--f).
In the subglacial channel, however,  the outputs can vary substantially for a given water discharge and width-integrated shear stress varies with respect to the other two  outputs as well (Figure~\ref{fig:Qw_vari} a--f).
For instance, very high  water velocity values and shear stresses can occur across minimal water discharge at $2.5$\,\unit{m}$^3$\,\unit{s}$^{-1}$ to the maximum water discharge at over $17$ \,\unit{m}$^3$\,\unit{s}$^{-1}$.
In \icesheet, mean values of water velocity can occur at water discharges between roughly $150$ \,\unit{m}$^3$\,\unit{s}$^{-1}$ and $310$ \,\unit{m}$^3$\,\unit{s}$^{-1}$.

When considering the subglacial channel's evolving width, width-integrated shear stress  across the channel generally increases with water discharge, with improved rank correlation compared to water velocity or shear stress (Figure~\ref{fig:Qw_vari} a--f).
Yet even the width-integrated shear stress  can vary substantially, with the highest values occurring at water discharge values ranging from roughly $11$ \,\unit{m}$^3$\,\unit{s}$^{-1}$ to over $17$ \,\unit{m}$^3$\,\unit{s}$^{-1}$.
The variability in width-integrated shear stress is less pronounced in the \icesheet case, with reduced variability in water discharge (Figure~\ref{fig:Qw_vari} c, f).
The persistence of high water velocity, shear stress, and width-integrated shear stress suggests that substantial sediment transport may occur across much of the water discharge range in the subglacial system (Figure~\ref{fig:Qw_vari} a--f).

@@

The second application examines the impact of a variety of channel shapes and friction factors on the relationship between water discharge and sediment transport capacity in both systems.
Across the range of parameter values examined, variability in water velocity and shear stress remains higher in the subglacial system compared to the subaerial one.
n \alpine{}, the variability in width-integrated shear stress is comparable between the two systems, due to the variable water discharge (Figure~\ref{fig:multi_run}).
 However, in \icesheet{}, with lower water discharge variability, the variability in width-integrated shear stress remains higher in the subglacial system across the time aggregation periods.

It is prescient to evaluate the time scales over which the subglacial channels may respond to changing water discharge to understand the influence of variability in sediment transport capacity.
In both cases, the variability in velocity, shear stress, and width-integrated shear stress decrease after $1-5$ day aggregations (Figure~\ref{fig:multi_run}).
Yet, substantial differences in variability persist in periods of up to $15$ days, as the variability in aggregations periods longer than this might be conflated with seasonal variations in water discharge.
Note that lags and time of peak sediment transport conditions still occur over these periods (Figure~\ref{fig:model_outs}), while the variability in sediment transport parameters decreases with larger time aggregation (Figure~\ref{fig:multi_run}).
\ian{add sentence here}

\subsection{Relationship between parameterization  outputs and observed sediment discharge}

When comparing the outputs with suspended sediment discharge measurements, the subaerial parameterizations generally show a better correlation with sediment discharge compared to subglacial outputs across the tested parameters (Figures~\ref{fig:Qw_vari} g--l and \ref{fig:multi_run}).
The rank correlation amongst the  subaerial parameterization's three output variables remains fairly constant despite the variable impact of channel shape and consistent scaling with water discharge (Figure \ref{fig:Qw_vari} and Section~\ref{sect:fluv}).
In subglacial channels, velocity and shear stress perform substantially worse in terms of rank correlation with sediment discharge compared to width-integrated shear stress (Figures~ \ref{fig:multi_run}).
In both \alpine{} and \icesheet{}, the  subglacial rank correlation approaches that of subaerial channels over longer time aggregation periods (Figures~ \ref{fig:multi_run} and \ref{fig:model_outs_1hr} to \ref{fig:model_outs_15day}).

\section{Discussion}
\subsection{Sediment capacity differences between the channel types}
\label{sect:dis_qsc}
Quantities that control sediment transport capacity, such as water velocity, shear stress, and width-integrated shear stress, vary differently in response to water discharge in subglacial channels compared to subaerial ones (Figure~\ref{fig:Qw_vari}).
In sediment transport relationships that evaluate sediment discharge capacity, such as in \citeA{meyer1948} or \citeA{engelund1967}, shear stress is scaled to the power of $\frac{3}{2}$ or $\frac{5}{2}$, respectively.
The exponent greater than $1$ magnifies sediment discharge variability beyond the variable sediment transport parameters described above (Figure~\ref{fig:multi_run}).
Note that sediment transport relationships describe sediment flux per unit width, so that the integrated capacity across channel occurs after application of the exponent, unlike the width-integrated shear stress evaluation here \cite<Equations\,\ref{eq:tautg} and \ref{eq:tautf}; c.f.>{tucker1997}.
Note also that both subglacial and subaerial parameterizations here assume that the distribution of velocity and shear stress is homogeneous across the channel bed, despite the limitations of this assumption \cite<e.g.>{yager2018}.

\ian{Mauro: Can you double-check this?}
The rapid increases and decreases in subglacial sediment transport capacity may compound the variability in sediment transport already present in subaerial systems \cite<Figure~\ref{fig:Qw_vari}; >{williams1989,jerolmack2010}.
By applying the discharge - width scaling in Equation~\ref{eq:wcf} to Equation~\ref{eq:v}, by setting $S\, \propto \,Q^{\frac{1}{2}}$, the subaerial channel's width-integrated shear stress response is $\tau_{tf}\, \propto\,  Q$ (see \ref{sect:scaling}). 
For sufficiently short time periods,  channel width and wetted area are fixed in the subglacial system so that the response of width-integrated shear stress is $ \tau_{tg}\, \propto\,  Q^{2}$, with a larger exponent on $Q$ than in the subaerial case following \citeA{alley1997}.
In this case, water discharge fluctuates at a shorter timescale than the subglacial channel evolution \cite{gimbert2016}.
Note also that the subglacial width integrated shear stress response to water discharge will be greater than the subaerial's provided that  Equation~\ref{eq:wcf}'s exponent is below one.
However, when subglacial channel size lies in equilibrium with water discharge, the  width-integrated shear stress response is $\tau_{tg}\, \propto \, Q^{\frac{4}{5}}$, similar to subaerial channels.
It follows that the stronger correlation between hydraulics and water discharge in \icesheet{} could result from being closer to this equilibrium between variations in the channels' wetter area and water discharge compared to \alpine{}.
This state could result from the reduced diurnal discharge amplitude in \icesheet{} (Figures~\ref{fig:model_outs} and ~\ref{fig:Qw_vari}).
In both cases, hydraulics could trend toward this equilibrium regime if the discharge is averaged discharge over $1-5$ days.
In this case, variability in subglacial channels  approaches that of subaerial channels (Figure~\ref{fig:multi_run}).
